Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-05-31 paper/msg.tex
Edits sections 4 and 5.

We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_{\overline{\partial} \Lambda_s} \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_{\overline{\partial} \Lambda_s}$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.

@@

Assume $\Omega \subset \RR^d$ is polygonal and that $\mathcal{T}$ is a triangulation of $\Omega$.  (If $d=1$ then instead $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
\begin{equation}
\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
\end{equation}
Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.

Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$.  From $b^h$ we define the closed and convex admissible subset of $\mathcal{V}^h$:
\begin{equation}
\mathcal{K}^h = \{r^h \in \mathcal{V}^h \,:\, r^h \ge b^h \text{ and } r^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
\end{equation}
For $s^h\in \mathcal{K}^h$ we also define the numerical icy domain as in \eqref{eq:lambdas}:
\begin{equation}
\Lambda_{s^h} = \{(x,y,z)\,:\,b^h(x,y) < z < s^h(x,y)\} \subset \RR^{d+1}.  \label{eq:felambdas}
\end{equation}
Also, a triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in the $s^h$ geometry if $s^h=b^h$ at every vertex of $T$, and icy otherwise.

Given $s^h$ and $b^h$ from $\mathcal{V}^h$, we construct a Firedrake extruded mesh \cite{McRaeetal2016} to decompose $\Lambda_{s^h}$.  If $d=2$ then the extruded mesh consists of triangular prisms, while if $d=1$ it has quadrilateral elements.  The extruded mesh construction starts from the triangulation $\mathcal{T}$ of $\Omega \subset \RR^d$, now called the ``base mesh''.  As shown in Figure \ref{fig:extruded}, each icy $T$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.  However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
\end{center}
\caption{We compute $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.  In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold); see \eqref{eq:ido}.  In the ``cliffs'' extrusion (right) all elements have a minimum thickness and the trace evaluation (bold) may occur above $s^h$ (dotted).}
\label{fig:extruded}
\end{figure}

We will compare two extrusion modes.  In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly, prism (quadilateral) elements at the ice margin degenerate; such elements have empty facets.  In the alternate ``cliffs'' extrusion, each element over an icy $T$ is nondegenerate.  Specifically, if $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in experiments, then each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$.  In this extrusion, a stress-free condition is applied on the cliffs  in the Glen-Stokes problem \eqref{eq:glenstokesweak}; they are treated the same way as the top surface.

Let $\Phi^h:\mathcal{K}^h \to (\mathcal{V}^h)'$ be the FE approximation of the IDO $\Phi$.  By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.  That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, but on $\overline{\partial} \Lambda_{s^h}$, and then extend by zero to all of $\Omega$.  In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.

We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}, and a Newton iteration, with direct solution of the step equations and back-tracking line search, to resolve the power-law nonlinearity.  As stated in Section \ref{sec:intro}, note that our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme (next section) are needed to solve the SIGP \eqref{eq:sigpweakform}.  Overall optimality of the solution algorithm would also need an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.

Finally, the discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
\begin{equation}
F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h , \label{eq:fesigpweakform}
\end{equation}
where $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.  We suppose that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin.

@@

In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations, variants of the Gauss-Seidel and Jacobi methods, which are suitable for VIs \cite{KinderlehrerStampacchia1980}.  While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.  However, because of the non-locality of the SIGP functional $F^h$, we must make careful approximations in order to retain adequate per-iteration performance.

Given a current surface elevation $s^h\in \mathcal{K}^h$, our smoothers will sweep through the $m$ vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems and then updating $s^h$.   Denote an arbitrary vertex of $\mathcal{T}$ by $(x_i,y_i)$ and let $\psi_i \in \mathcal{V}^h$ be the corresponding hat function, with $\psi_i(x_j,y_j)=\delta_{ij}$ \cite{Elmanetal2014}.  Also let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0\}$.  Note that $\psi_i \in \mathcal{D}_+^h$ and that $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  One can call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009}.

The one-dimensional VI at node $i$ is derived from \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for scalars $\gamma,c$.  Specifically, we seek $c \ge \beta_i^{s^h}$ such that
\begin{equation}
F^h(s_c)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \beta_i^{s^h}$.  Let
\begin{equation}
\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i} \label{eq:ferhoi}
\end{equation}
be the point-wise residual function and note \eqref{eq:fepointwiseviEARLY} simplifies into
\begin{equation}
(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
After solving \eqref{eq:fepointwisevi} for $c$ the surface elevation $s^h$ is then updated $s^h \gets s^h + c \psi_i$ by the smoother.  Observe that admissibility is preserved because the updated surface elevation is also in $\mathcal{K}^h$.

If $\rho_i(s^h; c)$ were linear and increasing in $c$ then the solution to \eqref{eq:fepointwisevi} would be given by a simple and well-known formula \cite[formula (4.4), for example]{GraeserKornhuber2009}, namely $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ where $\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$.  In fact $\rho_i(s^h; c)$ is nonlinear, but we adopt a simple approximation scheme for a smoother.  That is, we linearize
\begin{equation}
\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0) \label{eq:rhoapprox}
\end{equation}
and do one projected Newton step for problem \eqref{eq:fepointwisevi} at each node where $\rho_i'(s^h; 0) > 0$.  Because the interior PDE in the shallow, flat-bed model is elliptic, one can prove that the Jacobian diagonal entry in that case is positive on the icy part of $\Omega$ \cite{JouvetBueler2012}, but the full Glen-Stokes model here allows no such guarantee to our knowledge.  As the simplest-possible rule, if $\rho_i'(s^h; 0) \le 0$, so we know that the model has degenerated at the point in the sense that the linearized operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.  % FIXME THIS MUST BE TOO AGGRESSIVE?

Note that the derivative, which is also a Jacobian diagonal entry, can be approximated by a finite difference,
\begin{equation}
\rho_i'(s^h; 0) = (F^h)'(s^h)[\psi_i,\psi_i] \approx \frac{\rho_i(s^h; \eps) - \rho_i(s^h; 0)}{\eps}.  \label{eq:rhofd}
\end{equation}
Use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.

If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative.  By contrast, a (parallel, additive) Jacobi smoother would compute and store solutions $c$ at each node $i$ before doing any updates.  In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely the expense of yet more residual evaluations.  Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}, in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{pngs\_slow}(s^h,b^h,\id{eps}=1.0,\id{omega}=1.0)\text{:} \\+
    \ct{check admissibility: $s^h \ge b^h$} \\
    for $i = 0,\dots,m-1$ \\+
        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{FD for Jacobian diagonal entry} \\
        if $\alpha_i > 0$ \\+
            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
        else \\+
            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
\end{pseudo*}
\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.  Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of \pr{pngs\_slow}.}
\label{pc:pngsslow}
\end{pcode}

To build a better smoother we return to the computation of the Jacobian diagonal entry $\rho_i'(s^h; 0)$.  Suppose we partition the support of $\psi_i$ into where there is ice and not,
\begin{equation}
\theta_i = \{\psi_i > 0\} \cap \{s^h > b^h\}, \qquad {\hat\theta}_i = \{\psi_i > 0\} \setminus \theta_i.  \label{eq:thetasupport}
\end{equation}
Suppose, for the next two formulas, that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).  Then by \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute with local integrals
\begin{equation}
\rho_i(s^h; 0) = F^h(s^h)[\psi_i] - \ip{a}{\psi_i} = - \int_{\theta_i} (\bu|_{s^h} \cdot \bn_{s^h}- a)\, \psi_i  \label{eq:rhozero}
\end{equation}
and
\begin{align}
\rho_i'(s^h; 0) &= (F^h)'(s^h)[\psi_i,\psi_i]  \label{eq:rholocalderiv} \\
  &= - \int_{\theta_i} \lim_{\eps\to 0^+} \frac{(\bu|_{s^h+\eps\psi_i} - \bu|_{s^h}) \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i - \int_{{\hat\theta}_i} \lim_{\eps\to 0^+} \frac{\bu|_{s^h+\eps\psi_i} \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i.  \notag
\end{align}

Clearly, any smoother application, which is a sweep over all nodes in $\mathcal{T}$, will require at least one solution of the Glen-Stokes problem \eqref{eq:glenstokesweak}, namely solving with the geometry determined by the current surface elevation $s^h$.  However, formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated are prohibitively expensive because a Stokes solve is required  for each mesh node.  Thus we seek a less expensive way to estimate the surface velocity for a perturbed surface elevation with a small ``bump'' $\eps\psi_i$.  The change to the nonlinearities in this Glen-Stokes problem should be small, and it would seem that the most important change to the velocity and pressure fields from a small bump perturbation is through the perturbed gravitational load applied to the linearized problem around the state $s^h$.  That is, we suppose other effect such as a perturbed normal direction on the surface are significantly smaller.

@@

\renewcommand{\arraystretch}{1.1}
\begin{longtable}{l|l|l}
\toprule
\textbf{Acronym} {\Large$\strut$} & \textbf{Definition} & \textbf{Reference} \\ \hline
CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
FAS & full approximation scheme & Section \ref{sec:mcdstokes} \\
FE & finite element & Section \ref{sec:fe} \\
GS & Gauss-Seidel & Section \ref{sec:smoothers} \\
IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\
IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
MCDN & multilevel constraint decomposition (nonlinear) & Section \ref{sec:mcdstokes}, Pseudocode \ref{pc:mcdn-vcycle} \\
NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\
SIA & shallow ice approximation & Section \ref{sec:intro} \\
SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, equation \eqref{eq:strongform} \\
SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, equation \eqref{eq:ske} \\
VI & variational inequality & Section \ref{sec:weakido} \\
WU & work units & Section \ref{sec:mcdstokes} \\ % final \\ required
\bottomrule
\caption{Glossary of acronyms used in this paper.}
\label{tab:acronyms}
\end{longtable}

>>>>

We may now define the ice dynamics operator (IDO), a map which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}), as follows.  For each surface elevation function $s \in \mathcal{K}$ the IDO result $\Phi(s)$ is the normal component of the upper-surface trace of the unique velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the icy domain $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that the output $\Phi(s)$, a scalar function on $\Omega$, may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$, permitting substantial irregularities.

@@

Assume $\Omega \subset \RR^d$ is polygonal and that $\mathcal{T}$ is a triangulation of $\Omega$.  (If $d=1$ then $\mathcal{T}$ denotes an interval decomposition of $\Omega$.)  Based on low expected regularity at the ice margin, we will represent surface elevations $s\in \mathcal{K}$ using the $P_1$ finite element (FE) space
\begin{equation}
\mathcal{V}^h = \{s \in C^0(\Omega) : s|_T \text{ is linear if } T \in \mathcal{T}\} \subset W^{1,\qq}(\Omega).
\end{equation}
Note \cite{JouvetBueler2012} makes the same FE choice for the corresponding shallow problem.

Let $b^h \in \mathcal{V}^h$ be the discretized bed elevation, e.g.~the $\mathcal{V}^h$ interpolant of $b$, and define the following closed and convex admissible subset of $\mathcal{V}^h$:
\begin{equation}
\mathcal{K}^h = \{r^h \in \mathcal{V}^h \,:\, r^h \ge b^h \text{ and } r^h|_{\partial\Omega} = b^h|_{\partial\Omega}\}.  \label{eq:feK}
\end{equation}
Each $s^h\in \mathcal{K}^h$ defines a numerical icy domain $\Lambda_{s^h} \subset \RR^{d+1}$ as in \eqref{eq:lambdas}.  A triangle (interval) $T\in\mathcal{T}$ is said to be ice-free in $\Lambda_{s^h}$ if $s^h=b^h$ at every vertex of $T$, and icy otherwise.

Given $b^h \in \mathcal{V}^h$ and $s^h \in \mathcal{K}^h$, we use Firedrake to construct an extruded mesh \cite{McRaeetal2016} on $\Lambda_{s^h}$ consisting of triangular prisms (or quadrilaterals if $d=1$).  The triangulation $\mathcal{T}$ of $\Omega$ is now called the ``base mesh''.  As shown in Figure \ref{fig:extruded}, each icy $T \in \mathcal{T}$ generates a column of $m_z \ge 1$ prism (quadrilateral) elements in the extruded mesh.  However, ice-free $T$ have no extruded mesh elements at all; these columns are empty.  An extruded mesh is regenerated each time $F^h(s^h)$ is evaluated, but the base mesh $\mathcal{T}$ is reused.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/extruded.pdf}
\end{center}
\caption{We compute the IDO $\Phi^h(s^h)$ by solving \eqref{eq:glenstokesweak} on an extruded mesh with $m_z$ layers.  In the ``pinched'' extrusion (left), the trace of $\bu^h$ is evaluated on the graph of $s^h$ (bold).  In the ``cliffs'' extrusion (right) all elements have a minimum thickness so the trace evaluation may occur above $s^h$ (dotted).}
\label{fig:extruded}
\end{figure}

We will compare two extrusion modes.  In the ``pinched'' extrusion the elements cover $\Lambda_{s^h}$ exactly.  It follows that the elements at the ice margin are degenerate; they have positive volume but zero height at some nodes or even empty (vertical) facets.  In the alternate ``cliffs'' extrusion, each extruded element is nondegenerate.  Specifically, each prism (quadilateral) element has minimum thickness $H_{\text{min}}/m_z$ where $H_{\text{min}} > 0$ is the minimum total thickness, with $H_{\text{min}} = 20$ meters in computations.  The surface $s^h$ remains unchanged (and continuous), so these cliffs are only relevent to problem \eqref{eq:glenstokesweak}.

Let $\Phi^h:\mathcal{K}^h \to (\mathcal{V}^h)'$ be the FE approximation of the IDO $\Phi$.  By definition $\Phi^h(s^h)$ is the normal component of the numerical surface velocity after solving the Glen-Stokes equation \eqref{eq:glenstokesweak} on the domain $\Lambda_{s^h}$.  That is, we evaluate $\Phi^h(s^h)$ using the surface trace as in \eqref{eq:ido}, and then extend by zero to all of $\Omega$.  In the ``cliffs'' extrusion the trace is computed at the top of each column, and thus not necessarily at the original surface $s^h$.

We solve the Glen-Stokes problem \eqref{eq:glenstokesweak} using a $P^2 \times P^1$ Taylor-Hood mixed space for $(\bu,p)$ \cite{Elmanetal2014}.  The power-law nonlinearity is resolved by using a Newton iteration, with direct solution of the step equations and back-tracking line search.  As stated in Section \ref{sec:intro}, our goal in the current paper is to show that $O(1)$ iterations of the multilevel scheme are needed to solve the SIGP \eqref{eq:sigpweakform}.  Overall optimality of the solution algorithm also requires an optimal solution of the Glen-Stokes problem, such as the multigrid scheme in \cite{IsaacStadlerGhattas2015}.

Finally, let $F^h(s)[r] = - \ip{\Phi^h(s)}{r}$; compare \eqref{eq:sigpfunctional}.  The discrete SIGP computes $s^h \in \mathcal{K}^h$ satisfying
\begin{equation}
F^h(s^h)[r^h - s^h] \ge \ip{a}{r^h-s^h} \quad \text{for all } r^h \in \mathcal{K}^h. \label{eq:fesigpweakform}
\end{equation}
The only role of the extruded mesh in solving \eqref{eq:fesigpweakform} is in the evaluation of the IDO $\Phi^h(s^h)$.  We believe that \eqref{eq:fesigpweakform} is well-posed, but this is subject to the same caveats in the last section regarding identification of Sobolev spaces, along with the concern about overhangs at the margin, and we have no proof.

@@

In this section we propose solvers for \eqref{eq:fesigpweakform}, namely certain projected and nonlinear iterations which are variants of the Gauss-Seidel and Jacobi methods, but made suitable for VIs \cite{KinderlehrerStampacchia1980}.  Given a current surface elevation $s^h\in \mathcal{K}^h$, these iterations sweep through the $m$ interior vertices ($P_1$ nodes) of $\mathcal{T}$ in a fixed order, solving one-dimensional VI problems.  While these iterations are adequate smoothers in a multilevel method (next section), they converge slowly if used as single-level methods.  Because of the non-locality and density of the IDO $\Phi^h$ (Section \ref{sec:weakido}), we must make careful approximations in order to retain adequate per-iteration performance.

Denote an arbitrary vertex of $\mathcal{T}$ by $(x_j,y_j)$ and let $\psi_j \in \mathcal{V}^h$ be the corresponding $P_1$ hat function, with $\psi_j(x_k,y_k)=\delta_{jk}$ \cite{Elmanetal2014}.  Let $\mathcal{D}_+^h = \{t^h \in \mathcal{V}^h \,:\, t^h \ge 0 \text{ and } t^h|_{\partial\Omega} = 0\}$.  Let $i=0,\dots,m-1$ index the interior vertices of $\mathcal{T}$, so $(x_i,y_i) \notin \partial\Omega$.  Note that $\psi_i \in \mathcal{D}_+^h$, and $s^h + c \psi_i \in \mathcal{K}^h$ if and only if $c\ge \beta_i^{s^h} = b^h(x_i,y_i) - s^h(x_i,y_i)$.  We call $\beta_i^{s^h}$ the pointwise defect obstacle \cite{GraeserKornhuber2009} for $s^h$.

The one-dimensional VI at interior node $i$ is derived from the discrete SIGP \eqref{eq:fesigpweakform} using $r = s^h+\gamma \psi_i$ and $s = s^h+c \psi_i$ for some scalars $\gamma,c$.  Specifically, we seek $c \ge \beta_i^{s^h}$ such that
\begin{equation}
F^h(s^h+c \psi_i)[(s^h+\gamma \psi_i) - (s^h+c \psi_i)] \ge \ip{a}{(s^h+\gamma \psi_i) - (s^h+c \psi_i)} \label{eq:fepointwiseviEARLY}
\end{equation}
for all $\gamma \ge \beta_i^{s^h}$.  Let
\begin{equation}
\rho_i(s^h; c) = F^h(s^h+c\psi_i)[\psi_i] - \ip{a}{\psi_i} \label{eq:ferhoi}
\end{equation}
be the point-wise residual function, and observe that \eqref{eq:fepointwiseviEARLY} simplifies to
\begin{equation}
(\gamma - c) \,\rho_i(s^h; c) \ge 0. \label{eq:fepointwisevi}
\end{equation}
The smoother will approximately solve \eqref{eq:fepointwisevi} for $c$ the smoother, and then update $s^h \gets s^h + c \psi_i$.  Observe that admissibility is preserved because $s^h+c \psi_i \in \mathcal{K}^h$ when $c \ge \beta_i^{s^h}$.

If $\rho_i(s^h; c)$ were linear and increasing in $c$, e.g.~$\rho_i(s^h; c) = \rho_i(s^h; 0) + \alpha c$ with $\alpha > 0$, then the solution to \eqref{eq:fepointwisevi} would be given by the simple formula $c = \max\left\{-\rho_i(s^h; 0)/\alpha, \beta_i^{s^h}\right\}$ \cite{GraeserKornhuber2009}.  In fact $\rho_i(s^h; c)$ is nonlinear, and we will linearize
\begin{equation}
\rho_i(s^h; c) \approx \rho_i(s^h; 0) + c\, \rho_i'(s^h; 0), \label{eq:rhoapprox}
\end{equation}
and at each node where $\rho_i'(s^h; 0) > 0$ we will do one projected Newton step using the simple formula.  If $\rho_i'(s^h; 0) \le 0$, so that the model has degenerated in the sense that the linearized, pointwise operator is not acting elliptically with respect to a perturbation, we set $c = \beta_i^{s^h}$ to remove the ice at that location.  % FIXME TOO AGGRESSIVE?

Because the interior PDE in the shallow, flat-bed model is elliptic, it follows that $\rho_i'(s^h; 0) > 0$ on the icy part of $\Omega$ \cite{JouvetBueler2012}.  However, for the full Glen-Stokes model there is no such guarantee to our knowledge.

Note that the derivative $\rho_i'(s^h; 0)$, which is also a diagonal entry of the Jacobian of $F^h$, can be approximated by a finite difference,
\begin{equation}
\rho_i'(s^h; 0) = (F^h)'(s^h)[\psi_i,\psi_i] \approx \frac{\rho_i(s^h; \eps) - \rho_i(s^h; 0)}{\eps}.  \label{eq:rhofd}
\end{equation}
Literal use of \eqref{eq:rhofd} is very expensive because each value $\rho_i(s^h; \eps)$ requires a separate Glen-Stokes computation for the residual.

If each pointwise update $s^h \gets s^h + c \psi_i$ is done immediately then the smoother is of Gauss-Seidel (GS) type, also called serial or multiplicative, while a (parallel, additive) Jacobi smoother computes and stores solutions $c$ at every node $i$ before doing any updates.  In linear problems GS generally converges in about half as many iterations \cite{Greenbaum1997}, but here a more glaring difference dominates, namely that GS imposes the expense of yet more residual evaluations.  Nonetheless, the GS iteration using a single Newton step and approximation \eqref{eq:rhofd} is a well-defined smoother, so we state it in Pseudocode \ref{pc:pngsslow}.  It is given in a form which also allows over-relaxation ($\id{omega} > 1$) or under-relaxation ($\id{omega} < 1$) if desired.  Note that the Glen-Stokes problem \eqref{eq:glenstokesweak} is solved $2m$ times per application of this function.

\begin{pcode}[ht]
\begin{pseudo*}
\pr{pngs\_slow}(s^h,b^h,\id{eps}=1.0,\id{omega}=1.0)\text{:} \\+
    \ct{check admissibility: $s^h \ge b^h$} \\
    for $i = 0,\dots,m-1$ \\+
        $\alpha_i = (\rho_i(s^h; \eps) - \rho_i(s^h; 0))/\eps$  \qquad\qquad \ct{Jacobian diagonal entry} \\
        if $\alpha_i > 0$ \\+
            $c_i = - \rho_i(s^h; 0) / \alpha_i$ \\
            $(s^h)_i \gets \max\{(s^h)_i + \id{omega}\,c_i, (b^h)_i\}$ \\-
        else \\+
            $(s^h)_i \gets \beta_i$ \qquad\qquad \ct{non-elliptic case}
\end{pseudo*}
\caption{Projected nonlinear GS iteration, an expensive in-place smoother using a finite-difference (FD) derivative for the Jacobian diagonal.}
\label{pc:pngsslow}
\end{pcode}

To build a better smoother we return to the computation of the Jacobian diagonal entry $\rho_i'(s^h; 0)$.  Suppose we partition the support of $\psi_i$ into where there is ice and not,
\begin{equation}
\theta_i = \{\psi_i > 0\} \cap \{s^h > b^h\}, \qquad {\hat\theta}_i = \{\psi_i > 0\} \setminus \theta_i.  \label{eq:thetasupport}
\end{equation}
Recall that $\bu|_{z}$ denotes the surface velocity of the solution to the Glen-Stokes problem \eqref{eq:glenstokesweak} on the domain $\Lambda_{z}$, using an extruded mesh (Section \ref{sec:fe}).  By \eqref{eq:sigpfunctionalderiv}, \eqref{eq:idoderiv}, \eqref{eq:differencecases}, and \eqref{eq:ferhoi} we may compute
\begin{equation}
\rho_i(s^h; 0) = F^h(s^h)[\psi_i] - \ip{a}{\psi_i} = - \int_{\theta_i} (\bu|_{s^h} \cdot \bn_{s^h}- a)\, \psi_i  \label{eq:rhozero}
\end{equation}
and
\begin{align}
\rho_i'(s^h; 0) &= (F^h)'(s^h)[\psi_i,\psi_i]  \label{eq:rholocalderiv} \\
  &= - \int_{\theta_i} \lim_{\eps\to 0^+} \frac{(\bu|_{s^h+\eps\psi_i} - \bu|_{s^h}) \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i - \int_{{\hat\theta}_i} \lim_{\eps\to 0^+} \frac{\bu|_{s^h+\eps\psi_i} \cdot \bn_{s^h+\eps\psi_i}}{\eps} \psi_i.  \notag
\end{align}

Clearly, any smoother application, which is a sweep over all nodes in $\mathcal{T}$, will require at least one Glen-Stokes velocity solution, namely from solving \eqref{eq:glenstokesweak} on the geometry determined by the current surface elevation $s^h$.  Formulas \eqref{eq:rhozero} and \eqref{eq:rholocalderiv} as stated require a Stokes solution $\bu|_{s^h+\eps\psi_i}$ for each mesh node.

However, we may estimate the surface velocity for a such perturbed surface elevation, namely with a small ``bump'' $\eps\psi_i$, by replacing the geometrical bump with a local perturbation of the body force.  That is, we may perturb the gravitational load relative to the solution which gave $s^h$, but without changing the geometry.  The change to the nonlinearities in this Glen-Stokes problem should be small, and the most important change to the velocity and pressure fields from a small bump perturbation is through its weight.  We suppose that other effects, such as a perturbed normal direction on the surface, are significantly smaller.

@@

\renewcommand{\arraystretch}{1.1}
\begin{longtable}{l|l|l}
\toprule
\textbf{Acronym} {\Large$\strut$} & \textbf{Definition} & \textbf{Reference} \\ \hline
CMB & climatic mass balance & Section \ref{sec:stokesgeometry} \\
FAS & full approximation scheme & Section \ref{sec:mcdstokes} \\
FE & finite element & Section \ref{sec:fe} \\
GS & Gauss-Seidel & Section \ref{sec:smoothers} \\
IDO & ice dynamics operator & Section \ref{sec:weakido}, equation \eqref{eq:ido} \\
IIGP & implicit ice geometry problem & Section \ref{sec:evolution} \\
MCD & multilevel constraint decomposition & Section \ref{sec:mcdstokes} \\
MCDN & multilevel constraint decomposition (nonlinear) & Section \ref{sec:mcdstokes}, Pseudocode \ref{pc:mcdn-vcycle} \\
NCP & nonlinear complementarity problem & Section \ref{sec:stokesgeometry} \\
PNGS & projected, nonlinear Gauss-Seidel (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pngsslow} \\
PNJ & projected, nonlinear Jacobi (smoother) & Section \ref{sec:smoothers}, Pseudocode \ref{pc:pnj} \\
SIA & shallow ice approximation & Section \ref{sec:intro} \\
SIGP & steady ice geometry problem & Section \ref{sec:stokesgeometry}, equation \eqref{eq:strongform} \\
SKE & surface kinematical equation & Section \ref{sec:stokesgeometry}, equation \eqref{eq:ske} \\
VI & variational inequality & Section \ref{sec:weakido} \\
WU & work units & Section \ref{sec:mcdstokes} \\ % final \\ required
\bottomrule
\caption{Glossary of acronyms used in this paper.}
\label{tab:acronyms}
\end{longtable}
